HU Extension -- Assignment 02 -- E-185 Big Data Analytics 
========================================================
Handed out: 02/15/2013 -- Due by 17:25PM on Friday, 02/22/2013
-----------------------------------------------------------

#### **Problem 1.** Write an R function that would allow you to generate a matrix (data frame) with `m` columns and `n` rows and populate every column of that matrix with randomly generated values that fall between values `a` and `b`. Choose some arbitrary values for parameters a and b. Use your function to generate a matrix with 4 columns and a reasonably large value of the number of rows `n` , e.g. `1000-2000`. Plot histograms of data in two columns of your matrix and demonstrate to yourself that data are uniformly distributed, between selected values `a` and `b`. We are not performing any statistical tests for “uniformity” of the distribution. Just present data graphically in two different plots or histograms and take a look.

Selecting `a` as `-5` and `b` as `5`, we created the following function
```{r}
r_matrix <- function(n, m, a = -5, b = 5) {
  matrix(runif(n*m, min=a, max=b), nrow = n, ncol = m) 
}
```

Creating a matrix with 4 columns and 1500 rows

```{r}
X <- r_matrix(1500, 4)
```

Selecting 2 rows to plot the histograms

```{r fig.width=11, fig.height=6}
par(mfrow = c(1,2))
hist(X[,1], col = "light blue")
hist(X[,2], col = "light blue")
```

#### Problem 2. Try to find a way to present two distributions contained in any two of columns of your matrix on a single plot. To do that you might want to export the distribution data from two columns into two stand-alone vectors of equal length. Plot one distribution first using a call to `plot(x,y1)` function, where vector x contains the “predictor” or the parameter vector with values between `a` and `b` you selected above. To add the next curve (distribution `y2`) try invoking function `lines(x,y2)`. To improve your diagram, present two curves in different color and add labels on x and y axis, as well as the title to your graph. Try adding the distribution from the third column to your graph.

Defining the first column as the "predictor" and the following as y1, y2 and y3
```{r}
x <- X[1]
y1 <- X[2]
y2 <- X[3]
y3 <- X[4]
```